### 2015-09-15 manuscript.Rnw
edits to presentation of OTL's APIs

Generating phylogenies has been facilitated by advances in sequencing and
computing technologies enabling the investigation of evolutionary relationships
among taxa. In parallel, new phylogenetic comparative methods have been
developed and use these phylogenies to explore fundamental questions about the
origin of biodiversity including the evolution of morphological and ecological
traits, the spatio-temporal variation in speciation rates, or both
\cite{Pennell2013}. The development of these methods has contributed to the need
for a greater availability of high-quality phylogenetic information across
life. However, phylogenetic information is scattered, often only available as
image files within publications, and the variety of file formats used to store
phylogenetic information makes it difficult for researchers to access and
integrate into their own research (\cite{Stoltzfus2012}, but see
\cite{Cranston2014} for suggestions of best practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the tree to be updated continuously. The current draft of the Open Tree
of Life contains 2.3 million tips. Beyond obvious applications across the life
sciences to explore research questions in evolution, biodiversity, and
conservation, the resources OTL provides could be used in education and outreach
(e.g., illustrating course material, develop outreach activities to explore
relationships among species).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree
(\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
Application Programming Interface (API) to the different services it offers
allowing users to work programmatically with the underlying data. Here, we
present \texttt{rotl} an R package that allows users to interact with the OTL's
API to search for and import phylogenetic trees and taxonomic
information. Having access to this data directly from within R allows
researchers to take advantage of the multiple R packages that are available to
visualize, analyze and manipulate phylogenetic trees (ape, phytools, geiger,
ggtree, phylobase, etc... cite task view).

\section{Presentation of the services provided by OTL}

Presentation of the OTL: synthesis, 2.3M tips, continuously updated. There are
therefore 4 resources that the OTL provides: (1) the taxonomic information used
as the backbone from the tree, (2) the trees and associated published studies
that are used to build the synthetic tree, (3) a taxonomic name resolution
service used to map species names to the Open Tree Taxonomy, (4) the synthetic
tree. Also, Graph of Life endpoint but not of use for normal users and beyond
the scope of this paper.

\texttt{rotl} provides access to these services so the data stored in OTL
(including trees) can be imported into R (either in memory or by writing to
files) to interact with other packages and be used in comparative analyses, or
they can be written to files so they can used by other programs.

Interface with both the synthetic tree from Open Tree of Life and particular
trees from studies. Open Tree also features taxonomy lookup that can be used
with partial matching.

The OTL's API, and in turn \texttt{rotl}, provides services (1) to download a
tree from a set of species names, this can be useful to illustrate a manuscript,
map the evolution of a character in a taxonomic group, etc....; (2) to download
a tree published in a paper that might be used to build the synthetic tree, this
can be useful to reproduce or expand on an already published analysis; (3) to
download information about the taxonomy across all organisms considered by OTL
in a phylogenetic context.

Tree versionning? Replicability of analyses

Limitations

No branch lengths for trees from the synthetic TOL, limit applications for
comparative data, but could be possible to use calibrations from time tree
databases. Or combined with partial chronograms and use \cite{phyndr}.

>>>>

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies enabling the investigation of evolutionary
relationships among taxa. In parallel, new phylogenetic comparative methods have
been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \cite{Pennell2013}. The development of these methods has
contributed to the need for a greater availability of high-quality phylogenetic
information across life. However, phylogenetic information is scattered, often
only available as image files within publications, and the variety of file
formats used to store phylogenetic data makes it difficult for researchers to
access, synthesize, and integrate into their own research (\cite{Stoltzfus2012},
but see \cite{Cranston2014} for suggestions of best practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the tree to be updated as relationships are elucidated or
reevaluated. The current draft of the Open Tree of Life contains 2.3 million
tips. Beyond obvious applications across the life sciences to explore research
questions in evolution, biodiversity, and conservation, the resources OTL
provides could be used in education and outreach (e.g., illustrating course
material, develop outreach activities to explore relationships among species).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree
(\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
Application Programming Interface (API) to the different services it offers
allowing users to work programmatically with the underlying data. Here, we
present \texttt{rotl} an R package that allows users to interact with OTL's API
to search for, and import phylogenetic trees and taxonomic information into
R. Having access to this data directly in R allows researchers to take advantage
of the multiple R packages that are available to visualize, analyze and
manipulate phylogenetic and comparative data (ape, phytools, geiger, ggtree,
phylobase, RNeXML, etc... cite task view).

\section{Services provided by OTL}

The OTL project provides four resources that serve data to users through the API
(the API endpoints):

\begin{enumerate}
\item The taxonomic information used as the backbone from the tree, the Open
  Tree Taxonomy;
\item The studies and their associated trees some of which are chosen by
  curators to assemble the synthetic tree;
\item A taxonomic name resolution service (TNRS) used to map taxon names to
  the Open Tree Taxonomy;
\item The synthetic tree itself (the Open Tree).
\end{enumerate}

Additionally, the project provides access to other data, the Graph of Life, that
stores information about the underlying tree alignment graphs \cite{Smith2013}
used to synthesize the various sources of information from which the Open Tree
is assembled from. Data from this API endpoint is not intended for normal users
and its use will not be demonstrated in this paper but \texttt{rotl} has
functions to retrieve data it provides.

\texttt{rotl} gives users access to all endpoints provided by version 2 of the
OTL's API. Phylogenetic trees served by the API can be imported directly into
R's memory and are represented using the \texttt{ape} \cite{Paradis2004} tree
structure (objects of class \texttt{phylo}) or can be written to files in the
NEXUS, Newick or NeXML file formats. This allows researchers to use these trees
either directly with other R packages, or to be imported in other programs that
can read in phylogenetic tree files.

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used on the trees
returned by the API. However, resources and methods are being developed to add
branch lengths information to these topological trees (e.g., \cite{Ksepka2015},
\cite{Pennell2015}). Even without branch lengths, these trees could be useful to
illustrate relationships among species, or to map a trait on a phylogeny.

\texttt{rotl} can also be used to download trees accompanying studies, and that
are used to build the synthetic tree. All the trees from a study are available
even if they do not contribute to the synthetic tree. These trees provide a
useful resource to reproduce or expand on an already published analysis.
%% any examples available?

% Tree versionning? Replicability of analyses

\section{Technical information about \texttt{rotl}}

Use NCL/\texttt{rncl} to parse NEXUS/Newick files.

Taxonomic information, singletons, but NCL can parse it.
